A Stopped Negative Binomial Distribution

### 2016-02-29 stat-and-prob-letters/snb-paper.tex
Initial commit.

\begin{abstract}
This paper introduces a new discrete distribution suggested by curtailed
sampling rules common in early-stage clinical trials. We derive the
distribution of the smallest number of independent and identically
distributed Bernoulli trials needed in order to observe either $s$ successes 
or $t$ failures. The closed form expression for the distribution as well as 
the compound distribution are derived. Properties of the distribution are 
shown and discussed. A case study is presented showing how the distribution 
can be used to monitor sequential enrollment of clinical trials with binary 
outcomes as well as providing post-hoc analysis of completed trials.
\end{abstract}

@@

Consider the design for a Phase II clinical trial that investigates the
efficacy of iniparib in patients with breast cancer gene-associated (BRCA)
ovarian cancer [\ref{Sanofi 2013}]. In the first stage, 12 patients were
treated with iniparib (BSI-201/SAR240550). The outcome was defined by
Response Evaluation Criteria in Solid Tumor (RECIST). There were two
endpoints: terminate or enroll additional patients at the end of this
stage. If at least two of
these patients respond favorably, then the trial will proceed to the next
stage where additional patients were able to be treated. If fewer than two
respond then the trial would have been terminated.

The maximum sample size was 12 but the number of patients necessary to
reach any endpoint could be less. Our goal is to describe the distribution of
the enrollment size of the design for
planning purposes and to use this distribution to estimate the
success probability.  If all 12 patients are enrolled at once,
as in the classic
design, then the sample size is 12. However, in most clinical trials, the
patients are enrolled sequentially, often with one patient's outcome realized
before the next one enters the trial. In the present example, observing two
successful patients allows us reach one endpoint so the sample required
could be as small as two. So, 12 might not be necessary. Similarly 11
observed treatment failures also ends the stage. This sampling mechanism, in
which the experiment ends as soon as any of the endpoints is reached, is
call {\em curtailed sampling}. Under curtailed sampling the range of the
sample size is between two and 12.

We assume each of the patient outcomes can be modeled as independent,
identically distributed Bernoulli($p$) random variables. The realization of
the trial is a sequence of these random variables that stops when either a
specified number of success or failures is reached. In the
previous example suppose two successes were reached after enrolling 10
patients (one in the third step and one at the $10^{th}$). The sample
path is illustrated
in Fig.~\ref{fig:kane_viz}. The vertical axis denotes the number of
successful outcomes. The horizontal axis counts the number of patients that
have been enrolled. The horizontal and vertical boundaries represent
endpoints for the trial.

\begin{figure}[t!]
\includegraphics[width=\textwidth]{KanePlot.pdf}
\caption{
A hypothetical realization of the breast cancer trial.
}
\label{fig:kane_viz}
\end{figure}

The next section of this paper introduces our notation and basic results
including the density of the distribution along with a description of
it's relation to other distributions. Sections 2 and 3 derive the distribution
based on a defined Bernoulli process and give some basic properties of the
distribution. Section 4 derives the compound distribution using a Beta prior.
Section 5 develops a post-hoc analysis of a completed trial. Section 6 is
devoted to discussion and conclusions. The simulations, model fitting
routines, and visualizations presented in this paper were generated using the
snb package [\ref{Kane 2015}] for the R programming
environment [\ref{R Core Team 2015}]

\section{Notation and Summary of Distributional Results}
\label{notation.section}

@@

The distribution of $\,Y\,$ has support on integer values in the range
\begin{equation*}               
     \min(s,t) \leq \; Y \;\leq s+t-1  \label{range.y.eq}
\end{equation*}
and it is distributed as
\begin{equation} \label{eqn:pmf}
\mathbb{E}\  I_{\{Y=k\}} = S(k, p, s) \ I_{\{s \leq k\}} + 
  S(k, 1-p, t) \ I_{\{ t \leq k \}}
\end{equation}
where $I_{\{f\}}$ is one if $f$ is true and zero otherwise and
\begin{equation} \label{eqn:N}
S(k, p, s) = {k-1 \choose s-1} p^s (1-p)^{k-s} 
\end{equation}
$S$ is the negative binomial probability
In~\ref{eqn:pmf} we can see that the mass at
any point on the support of the distribution may come from $S(p)$; $S(1-p)$;
or both of their probabilities.

To show the result in Eqn.~\ref{eqn:pmf} consider the
process $\mathbf{X} = \left\{X(k) : k = 0,1,... \right\}$
with $X(0)=0$ and
\begin{equation*} \label{eqn:proc}
X_{k+1} = X_k + b_{k+1} \ I_{\{ k-t < X_k < s\}}.
\end{equation*}
The process can be conceptualized as a series of coin
flips that stops when either $s$ successes or $t$ failures are reached.
At each step a coin is flipped. If it is heads, the process advances
one diagonally in the
positive horizontal and vertical direction in Fig.~\ref{fig:kane_viz}.
Otherwise, it advances in the positive horizontal direction only. When the
process stops either $X_k = s$ or $X_k = k-t$.

\begin{prop}
The distribution of the stopping time
$\argminl_k \left\{ X_k \geq s \cup X_k \leq k-t \right\}$
is equal to the SNB and is given in (\ref{eqn:pmf}).
\end{prop}
\begin{proof}
%The proof will proceed in two parts. First, a combinatorial justification 
%will be given for the probability mass value on each element of the support. 
%Second, it will be shown that the sum of the masses over the support sums to 
%one.

The probability that a given realization of $\mathbf{X}$ reaches $s$ at
the $k$th outcome is the probability that, at time $k-1$ there are $s-1$
successful outcomes and $k-s$ unsuccessful outcomes multiplied by
the probability of a
success at time $k$ (\ref{eqn:N}). The probability $X_k = k-t$
is the probability that, at outcome $k-1$ there are $k-t$ successful outcomes
and $t-1$ unsuccessful outcomes multiplied by the probability of an
unsuccessful outcome at time $k$.  Finally, we need to show that
the sum
\begin{align} \label{eqn:sum_proof}
R &= \sum_{k=s}^{s+t-1} S(k, p, s) + \sum_{k=t}^{s+t-1} S(k, 1-p, t) \\
  &= \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s (1-p)^{k-s} + \sum_{k=t}^{s+t-1} {k-1 \choose k-t} p^{k-t} (1-p)^t
\end{align}
is equal to one.

Substitute $i=k-s$ in the first summation and
$j=k-t$ in the second.
Then $R$ can be written
as the cumulative distribution function of two
negative binomial distributions
\begin{equation} \label{eqn:transformed_sum}
R = \sum_{i=0}^{t-1} {i+s-1 \choose i} p^s (1-p)^i \; + \;
\sum_{j=0}^{s-1} {j+t-1 \choose j} p^j (1-p)^t.
\end{equation}

Let $\mathcal{I}_p(s, t)$ be the {\em regularized incomplete beta function}, which is
also one minus the c.d.f. of the negative binomial distribution. This
function satisfies $\mathcal{I}_p(s, t) = 1-\mathcal{I}_{1-p}(t, s)$
\citep{Abramowitz1964}.
Then it can be proven
that \ref{eqn:pmf} is a valid probability mass by showing that the
probability mass function sums to one.
\begin{align*}
R = \sum_{i=0}^{t-1} &{i+s-1 \choose i} p^s (1-p)^i +
\sum_{j=0}^{s-1}  {j+t-1 \choose j} p^j  (1-p)^t \\
   &= 1-\mathcal{I}_p(s, t) + 1 - \mathcal{I}_{1-p}(t, s) \\
   &= 1. 
\end{align*}
\end{proof}

\begin{prop} Let $S$ be distributed SNB with parameters $p$, $s$, and $t$.
Then the moment generating function (MGF) of $S$ is
\begin{equation} \label{eqn:mgf}
\mathbb{E}~e^{xS} = \frac{p e^x}{1 - qe^x} \mathcal{I}_{qe^x} (s, t) + 
  \frac{qe^x}{1-pe^x} \mathcal{I}_{pe^x}(s, t)
\end{equation}
for $q = 1-p$ when $x \leq -\log(q)$ and $x \leq -\log(p)$.
\end{prop}
\begin{proof}
By definition, the MGF of the SNB is:
\begin{equation*}
\mathbb{E}~e^{xS} = \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s q^{k-s} e^{kx} 
  + \sum_{k=t}^{s+t-1} {k-1 \choose k-t} p^{k-t} q^t e^{kx}
\end{equation*}
and this can be rewritten as:
\begin{equation} \label{eqn:first_sum}
\mathbb{E}~e^{xS} = \sum_{k=s}^{s+t-1}{k-1 \choose s-1} (pe)^{sx} (qe^x)^{k-s} 
  + \sum_{k=t}^{s+t-1}{k-1 \choose k-t} (qe^x)^t (pe^x)^{k-t}.
\end{equation}
Taking the first summation in Equation \ref{eqn:first_sum}:
\begin{align*}
\sum_{k=s}^{s+t-1}{k-1 \choose s-1} (pe)^{sx} (qe^x)^{k-s} &= 
  \left(\frac{pe^x}{1 - qe^x}\right)^s \ \ \sum_{k=s}^{s+t-1} {k-1 \choose s-1} 
    (qe^x)^{k-s} (1-qe^x)^s) \\
  &= \left(\frac{pe^x}{1 - qe^x}\right)^s \mathcal{I}_{qe^x}(s, t).
\end{align*}
Since the incomplete beta function has support on zero to one
$qe^x \leq 1$. This implies that $x \leq -\log(q)$.

@@

The probability mass function of $\,Y\,$ has a variety of shapes for different choices of the parameters $(s,\, t,\,p)$.  These shapes are illustrated in 
Fig.~\ref{shapes.fig}.The SNB is related to the negative binomial 
distribution. Specifically, if $t$ is large then the $Y-s$ has a negative 
binomial distribution with
\begin{equation*}                                    %   (1)   
\mathbb{P}\{Y=s+j \}        \label{nb1.eq}          
  = {{s+j-1}\choose{s-1}} p^s (1-p)^j
\end{equation*}
for $\,j=0, 1,\ldots\,$. A similar statement can be made when $s$ is large
and $t$ is small.

For the special case of $\,s=t,$ the distribution of $\,Y\,$ is the
riff-shuffle, or minimum negative binomial distribution~\citep{Uppuluri1970}.
Similar derivations of the closely-related maximum negative binomial 
discretedistributions also appear in~\cite{Zhang 2000}
and \cite{Zelterman2004}.
The maximum negative binomial is the smallest number of outcomes necessary to 
observe at least $c$ successes {\em and} $c$ failures, but the SNB givethe 
number of coin flips to observe {\em either} $s$ heads or $t$ tails.

@@

Eqn.~\ref{eqn:posterior} suggests an alternative construction for
Phase II clinical trials.  Patients could be
enrolled sequentially until there are enough patients to guarantee the
desired level of power or until a specified number of adverse events are
recorded. In cases such as these the endpoints $s$ and $t$ are known and
new outcomes can be incorporated into the estimate of
$p$ to provide updated odds of successful enrollment. Later it will be shown
that estimates of the distribution of $p$ can even be used after an endpoint
has been reached to determine the amount of uncertainty in the endpoint.

>>>>

\begin{abstract}
This paper introduces a novel discrete distribution suggested by curtailed
sampling rules common in early-stage clinical trials. We derive the
distribution of the smallest number of independent and identically
distributed Bernoulli trials needed in order to observe either $s$ successes 
or $t$ failures. The closed-form expression for the distribution as well as 
its characteristics and properties are explored.
\end{abstract}

@@

Consider a prototypical Phase II single-arm clinical trial. 12 patients
are enrolled and treated. If at least 2 of those patients respond
favorably, then the trial will proceed to the next stage. If fewer than 
two respond then the trial would have been terminated.

The maximum sample size is 12 but the number of patients necessary to
reach any endpoint could be less. Our goal is to describe the distribution of
the enrollment size. This distribution will be used for planning purposes. 
If all 12 patients are enrolled at once, as in the classic
design, then the sample size is 12. However, in most clinical trials, the
patients are enrolled sequentially, often with one patient's outcome realized
before the next one enters the trial. In the present example, observing two
successful patients allows us reach one endpoint so the sample required
could be as small as two - 12 might not be necessary. Similarly 11
observed treatment failures also ends the stage. This sampling mechanism, in
which the experiment ends as soon as any of the endpoints is reached, is
call {\em curtailed sampling}. Under curtailed sampling the range of the
sample size is between two and 12.

Assume each of patient outcome can be modeled as an independent,
identically distributed Bernoulli($p$) random variable. The trial is realized
as a sequence of these random variables that stops when either a
specified number of success or failures is reached. In the
previous example suppose two successes were reached after enrolling 10
patients (one in the third step and one at the $10^{th}$). The sample
path is illustrated
in Fig.~\ref{fig:kane_viz}. The vertical axis denotes the number of
successful outcomes. The horizontal axis counts the number of patients that
have been enrolled. The horizontal and vertical boundaries represent
endpoints for the trial.

\begin{figure}[t!]
\includegraphics[width=\textwidth]{KanePlot.pdf}
\caption{
A hypothetical realization of a trial.
}
\label{fig:kane_viz}
\end{figure}

The rest of this paper derives the distribution for the number of
patients needed to end a trial and explores the characteristics and
properties of this distribution.
The next section introduces our notation and basic results
including the density of the distribution along with a description of
it's relation to other distributions. Sections 2 derives the distribution
based on a defined Bernoulli process and give some basic properties.
Section 3 provides a connection to the Binomail tail probability.
Section 4 derives the posterior distribution using a Beta prior.
Section 5 provides a brief discussion on the use of the distribution
in clinical trials along with future avenues for generalization.

\section{Distributional Result}
\label{notation.section}

@@

The distribution of $\,Y\,$ has support on integer values in the range
\begin{equation*}               
     \min(s,t) \leq \; Y \;\leq s+t-1  \label{range.y.eq}
\end{equation*}
and it is distributed as
\begin{equation} \label{eqn:pmf}
\mathbb{E}\  I_{\{Y=k\}} = S(k, p, s) \ I_{\{s \leq k\}} + 
  S(k, 1-p, t) \ I_{\{ t \leq k \}}
\end{equation}
where $I_{\{f\}}$ is one if $f$ is true and zero otherwise and
\begin{equation} \label{eqn:N}
S(k, p, s) = {k-1 \choose s-1} p^s (1-p)^{k-s} 
\end{equation}
$S$ is the negative binomial probability

To prove the result in Eqn.~\ref{eqn:pmf} consider the
process $\mathbf{X} = \left\{X(k) : k = 0,1,... \right\}$
with $X(0)=0$ and
\begin{equation*} \label{eqn:proc}
X_{k+1} = X_k + b_{k+1} \ I_{\{ k-t < X_k < s\}}.
\end{equation*}
The process can be conceptualized as a series of coin
flips that stops when either $s$ successes or $t$ failures are reached.
At each step a coin is flipped. If it is heads, the process advances
one diagonally in the
positive horizontal and vertical direction.
Otherwise, it advances in the positive horizontal direction only. The
process stops when either $X_k = s$ or $X_k = k-t$.

\begin{prop}
The distribution of the stopping time
$\argminl_k \left\{ X_k \geq s \cup X_k \leq k-t \right\}$
is equal to Eqn.~\ref{eqn:pmf}.
\end{prop}
\begin{proof}
%The proof will proceed in two parts. First, a combinatorial justification 
%will be given for the probability mass value on each element of the support. 
%Second, it will be shown that the sum of the masses over the support sums to 
%one.

The probability a given realization of $\mathbf{X}$ reaches $s$ at
the $k$th outcome is the probability that, at time $k-1$ there are $s-1$
successful outcomes and $k-s$ unsuccessful outcomes multiplied by
the probability of a
success at time $k$. 
\begin{equation} \label{stop_s}
S(k, p, s) = {k-1 \choose s-1} p^s (1-p)^{k-s}
\end{equation}
The probability a given realization reaches $k-t$
is the probability that, at outcome $k-1$ there are $k-t$ successful outcomes
and $t-1$ unsuccessful outcomes multiplied by the probability of an
unsuccessful outcome at time $k$.  
\begin{equation} \label{stop_t}
S'(k, p, t) = {k-1 \choose k-t} p^{k-t} (1-p)^t
\end{equation}
It can be shown that $S(k, p, s) = S'(k, 1-p, s)$ by realizing that 
${k-1 \choose k-s} = {k-1 \choose s-1}$.

To show that sum of Eqn. \ref{stop_s} and \ref{stop_t} sum to one
over their support 
\begin{align} \label{eqn:sum_proof}
R &= \sum_{k=s}^{s+t-1} S(k, p, s) + \sum_{k=t}^{s+t-1} S(k, 1-p, t) \\
  &= \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s (1-p)^{k-s} + \sum_{k=t}^{s+t-1} {k-1 \choose k-t} p^{k-t} (1-p)^t
\end{align}
substitute $i=k-s$ in the first summation and $j=k-t$ in the second.
Then $R$ can be written
as the cumulative distribution function of two
negative binomial distributions
\begin{equation} \label{eqn:transformed_sum}
R = \sum_{i=0}^{t-1} {i+s-1 \choose i} p^s (1-p)^i \; + \;
\sum_{j=0}^{s-1} {j+t-1 \choose j} p^j (1-p)^t.
\end{equation}

Let $\mathcal{I}_p(s, t)$ be the {\em regularized incomplete beta function}
and recall this function satisfies 
$\mathcal{I}_p(s, t) = 1-\mathcal{I}_{1-p}(t, s)$ \citep{Abramowitz1964}.
\begin{align*}
R = \sum_{i=0}^{t-1} &{i+s-1 \choose i} p^s (1-p)^i +
\sum_{j=0}^{s-1}  {j+t-1 \choose j} p^j  (1-p)^t \\
   &= 1-\mathcal{I}_p(s, t) + 1 - \mathcal{I}_{1-p}(t, s) \\
   &= 1. 
\end{align*}
\end{proof}

\section{Shape and Basic Properties}

The probability mass function of $\,Y\,$ has a variety of shapes for different 
choices of the parameters $(s,\, t,\,p)$.  These shapes are illustrated in 
Fig.~\ref{shapes.fig}. The SNB is related to the negative binomial 
distribution. Specifically, if $t$ is large then the $Y-s$ has a negative 
binomial distribution with
\begin{equation*}                                    %   (1)   
\mathbb{P}[Y=s+j]        \label{nb1.eq}          
  = {{s+j-1}\choose{s-1}} p^s (1-p)^j
\end{equation*}
for $\,j=0, 1,\ldots\,$. A similar statement can be made when $s$ is large
and $t$ is small.

For the special case of $\,s=t,$ the distribution of $\,Y\,$ is the
riff-shuffle, or minimum negative binomial distribution~\citep{Uppuluri1970}.
Similar derivations of the closely-related maximum negative binomial 
discrete distributions also appear in~\cite{Zhang2000}
and \cite{Zelterman2005}.
The maximum negative binomial is the smallest number of outcomes necessary to 
observe at least $c$ successes {\em and} $c$ failures, but the SNB givethe 
number of coin flips to observe {\em either} $s$ heads or $t$ tails.

\begin{prop} Let $Y$ be distributed as SNB($p$, $s$, $t$) and let 
$B$ be distributed Binomial with size $n=s+t-1$ and success probability
$p$. Then
\begin{equation}
\mathbb{P}[B \geq s] = \mathbb{P} [Y \leq n | \text{\#success} = s].
\end{equation}
That is, the probability that the number of successes is at least $s$
in the Binomial model is the same that the trial stops with $s$ 
successes in the SNB model.
\end{prop}
\begin{proof}
The Binomial tail probability is
\begin{align*}
\mathbb{P}[B \geq s] &= \sum_{k=s}^{s+t-1} {n \choose k} p^k (1-p)^{n-k} \\
  &= 1 - \sum_{k=0}^{s-1} p^k (1-p)^{n-k} \\
  & 1 - \mathcal{I}_{1-p}.
\end{align*}
Next, consider the SNB probability
\begin{equation*}
\mathbb{P} [Y \leq n | \text{\#success} = s] = \sum_{k=s}^{s+t-1}
  {k-1 \choose s-1} p^s (1-p)^k-s.
\end{equation*}
Let $i=k-s$ then using the fact that ${i+s-1 \choose s-1} = {i+s-1 \choose i}$
the summation can be rewritten as
\begin{align}
\mathbb{P} [Y \leq n | \text{\#success} = s] &= \sum_{i=0}^{t-1} 
  {i+s-1 \choose i} p^s (1-p)^i\\
  &= 1 - \mathcal{I}_{1-p}(t, s).
\end{align}
\end{proof}

\section{The Moment Generating Function}

\begin{prop} Let $S$ be distributed SNB with parameters $p$, $s$, and $t$.
Then the moment generating function (MGF) of $S$ is
\begin{equation} \label{eqn:mgf}
\mathbb{E}~e^{xS} = \frac{p e^x}{1 - qe^x} \mathcal{I}_{qe^x} (s, t) + 
  \frac{qe^x}{1-pe^x} \mathcal{I}_{pe^x}(s, t)
\end{equation}
for $q = 1-p$ when $x \leq -\log(q)$ and $x \leq -\log(p)$.
\end{prop}
\begin{proof}
By definition, the MGF of the SNB is:
\begin{equation*}
\mathbb{E}~e^{xS} = \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s q^{k-s} e^{kx} 
  + \sum_{k=t}^{s+t-1} {k-1 \choose k-t} p^{k-t} q^t e^{kx}
\end{equation*}
and can be rewritten as:
\begin{equation} \label{eqn:first_sum}
\mathbb{E}~e^{xS} = \sum_{k=s}^{s+t-1}{k-1 \choose s-1} (pe)^{sx} (qe^x)^{k-s} 
  + \sum_{k=t}^{s+t-1}{k-1 \choose k-t} (qe^x)^t (pe^x)^{k-t}.
\end{equation}
Taking the first summation in Equation \ref{eqn:first_sum}:
\begin{align*}
\sum_{k=s}^{s+t-1}{k-1 \choose s-1} (pe)^{sx} (qe^x)^{k-s} &= 
  \left(\frac{pe^x}{1 - qe^x}\right)^s \ \ \sum_{k=s}^{s+t-1} {k-1 \choose s-1} 
    (qe^x)^{k-s} (1-qe^x)^s) \\
  &= \left(\frac{pe^x}{1 - qe^x}\right)^s \mathcal{I}_{qe^x}(s, t).
\end{align*}
Since the incomplete beta function has support on zero to one
$qe^x \leq 1$. This implies that $x \leq -\log(q)$.

@@

@@
